---
title: "P8451 Machine Learning in Public Health - Assignment 1"
output: github_document
date: "2023-2-14"
---

In preparation for the analyses below, we will load the following libraries:
```{r}
library(tidyverse)
library(dplyr)
```

# Part I: Implementing a Simple Prediction Pipeline

## Question 1: Fitting Prediction Models

### 1.1 Data Import and Cleaning 

We will begin by importing the general health and physical activity data collected by the __New York City Department of Health__ using the `read_csv` function. 

```{r}
nyc_health_data = read_csv("./class4_p1.csv") %>% 
  janitor::clean_names() %>% 
  mutate(id = x1, 
         hypertension = factor(chronic1, labels = c("Yes", "No")),
         diabetes = factor(chronic3, labels = c("Yes, No")), 
         asthma = factor(chronic4, labels = c("Yes", "No")), 
         smoking = factor(tobacco1, labels = c("Most or All Days", 
                                               "Some Days", 
                                               "Never")), 
         alcohol = factor(alcohol1, labels = c("Most or All Days", 
                                               "Some Days", 
                                               "Never")), 
         physical_activity_minutes = gpaq8totmin, 
         walk_days = gpaq11days, 
         physical_activity_cat = factor(habits5, labels = c("Very Active", 
                                                            "Somewhat Active", 
                                                            "Not Very Active", 
                                                            "Not Active At All")), 
         diet_cat = factor(habits7, labels = c("Excellent",
                                               "Very Good", 
                                               "Good", 
                                               "Fair", 
                                               "Poor")), 
         age_cat = factor(agegroup, labels = c("18-24 Yrs", 
                                               "25-44 Yrs", 
                                               "45-64 Yrs", 
                                               "65+")), 
         sex = factor(dem3, labels = c("Male", "Female")), 
         hispanic = factor(dem4, labels = c("Yes", "No")), 
         born_in_US = factor(dem8, labels = c("USA", "Outside USA")), 
         poverty_group = factor(povertygroup, labels = c("<100%", 
                                                         "100-199%", 
                                                         "200-399%", 
                                                         "400-599%", 
                                                         "600% +", 
                                                         "Don't Know")), 
         healthy_days = healthydays) %>% 
  select(id, hypertension, diabetes, asthma, bmi, smoking, alcohol, 
         physical_activity_minutes, walk_days, physical_activity_cat, diet_cat, 
         age_cat, sex, hispanic, born_in_US, poverty_group, healthy_days) %>% 
  na.omit() %>% 
  distinct()
```

### 1.2 Data Import and Cleaning 